10 years of school closures and consolidations: the impact on school accessibility

### 2024-12-12 paper/Manuscript.qmd
draft revised manuscript - for other author reviews

---
title: "School closures and consolidations: the active travel and emission implications of reduced accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto

author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing emissions and increasing active travel are top priorities for many communities. Often, though, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools. Public schools have been strained to do more with less, and governments are opting to close under-capacity schools and expand existing schools to accommodate lost capacity. The objective of this study is to quantify the travel burdens of school consolidation/closure policies. We do this through the case of Hamilton, a mid-sized city in Ontario, Canada. Between the years 2011 and 2015, 7% of the elementary schools closed in the city. We study how these closures changed the accessibility landscape for the elementary school aged population, as well as the active travel and emission implications. In terms of methods, we use spatial availability, a singly-constrained multimodal measure of spatial accessibility. Thanks to its proportional allocation mechanism, spatial availability allows for an interpretable comparison between 'before' and 'after' the implementation of school closure/consolidation policy. Analysis is conducted at the parcel-level with all school seats being eligible and travel behaviour based on observed home-to-school patterns for motorized and non-motorized modes. We report estimates of system-wide impacts that result from the closures, such as potential growth in GHG emission and reduction of walking minutes. Findings demonstrate that overall, spatial availability decreased in the city. Though the school consolidation policy may have resulted in cost-savings in operation and maintenance of facilities, students, families, and communities pay the price in reduced opportunities for physical activity and greater motorized mobility-related pollution.
keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---
<!--NOTE WHEN PREPARING THE .TEX for journal submission I ran into errors as their system uses PDFlatex+Bibtex to compile. The following packages only work in lualatex, so do the following:
- remove "  \usepackage[T1]{fontenc}" 
- remove "\usepackage{fontspec}" 
- add "\DeclareUnicodeCharacter{394}{$\Delta$}" to the tex preamble (this is because PHDtex does not recognize unicode? So need to explicitly define) (add to line 5) -->

<!--Chat with Antonio about the assumption regarding the 'student population' per parcel. As it stands, when the student pop for the OD matrix is summed, it equals the total student population in the city. Each parcel is assigned a DA-pop/(the number of potential parcel trips in a DA * the potential parcels trips from that parcel). This is highly correlated with we just weighted it by DA-pop/DA-dwellings (but need to double check).  --->

@@

```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
library(miscTools) # Miscellaneous Tools and Utilities
library(huxtable)
library(tiff)
library(grid)
```

@@

ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"),
                          quiet = TRUE) |> st_transform(st_crs(cen_data_2016_hamiltononly))
ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),quiet = TRUE) |> 
  st_transform(st_crs(cen_data_2016_hamiltononly)) |> 
  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                               COMMUNITY_ == "Dundas" ~ "DU",
                               COMMUNITY_ == "Flamborough" ~ "FL",
                               COMMUNITY_ == "Glanbrook" ~ "GL",
                               COMMUNITY_ == "Hamilton" ~ "HA",
                               COMMUNITY_ == "Stoney Creek" ~ "SC"),
         COMMUNITY_ = ifelse(COMMUNITY_ == "Hamilton", "Hamilton Central", COMMUNITY_)) |> st_transform(st_crs(cen_data_2016_hamiltononly))

@@

cen_data_2021_hamiltononly <- 
  cen_data_2021_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA21_566: Median after-tax income in 2020 among recipients ($)" ,"v_CA21_71: 15 to 19 years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA21_966: Median after-tax income of economic family in 2020 ($)",
    "DA_med_AT_lone-family_income" = "v_CA21_978: Median after-tax income of one-parent economic families in 2020 ($)",
    "DA_prev_LIMAT_under18" =  "v_CA21_1043: 0 to 17 years",
    "DA_pop_5to9" = "v_CA21_32: 5 to 9 years",
    "DA_pop_10to14" = "v_CA21_50: 10 to 14 years") %>%
    mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))
```

@@

```{r augmenting-catchment, include=FALSE}
# ALL_SCHOOLS_CATCH_Elem_extra2011 <- ALL_SCHOOLS_CATCH_Elem |>
#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2011) |> 
#               summarize(OTGC = sum(OTGC2011, na.rm = T)), 
#             by = c("CID" = "CID2011")) |> filter(!is.na(OTGC))
# 
# ALL_SCHOOLS_CATCH_Elem_extra2016 <- ALL_SCHOOLS_CATCH_Elem |>
#   left_join(ALL_SCHOOLS_Elem |> st_drop_geometry() |> group_by(CID2016) |> 
#               summarize(OTGC = sum(OTGC2016, na.rm = T)), 
#             by = c("CID" = "CID2016")) |> filter(!is.na(OTGC))
# 
# ALL_SCHOOLS_CATCH_Elem_extra <- rbind(ALL_SCHOOLS_CATCH_Elem_extra2011, ALL_SCHOOLS_CATCH_Elem_extra2016)
# rm(ALL_SCHOOLS_CATCH_Elem_extra2011,ALL_SCHOOLS_CATCH_Elem_extra2016)
# 
# #checks
# sum(ALL_SCHOOLS_Elem$OTGC2011, na.rm=T) + sum(ALL_SCHOOLS_Elem$OTGC2016, na.rm=T)
# ALL_SCHOOLS_CATCH_Elem_extra$OTGC |> sum(na.rm=T)

#reformatting schools for mapping
SCHOOLS_nochangeorexpanded <- SCHOOLS |> filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021") |> 
  mutate(Year = ifelse(Year == "2011 and 2016", "2011", Year))
ALL_SCHOOLS_Elem_formapping <- rbind(SCHOOLS, SCHOOLS_nochangeorexpanded)
rm(SCHOOLS_nochangeorexpanded)
ALL_SCHOOLS_Elem_formapping <- ALL_SCHOOLS_Elem_formapping |> mutate(Year = ifelse(Year == "2011 and 2016", "2016", Year))
```

```{r descriptive-catchment-schools-plot, eval=FALSE}
tmap_defaults <-
  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
  tm_compass(size = 1.5, position = c("right", 'top')) +
  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,

catchment_schools_plot <-
   tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(ALL_SCHOOLS_CATCH_Elem_extra) +
  tm_polygons(col = "OTGC", border.alpha=0.1, palette = "RdPu",
              labels = c("[181-318)", "[318-406)", "[406-566)","[566-1417]"),
              breaks = c(181, 318, 406, 566, 1417),
              title = "On the ground\ncapacity (OTGC)")+
  tm_facets(by=c("System","Year"))+
  tm_shape(ALL_SCHOOLS_Elem_formapping) +
  tm_symbols(shape = "Status",
             size=0.15,
          title.shape = "School status",
          shapes = c(24,25,21,23,22),
          shapes.labels = c(
          "Expanded after 2011",
          "Moved before 2016",
          "New after 2011",
          "No change btwn 2011-16",
          "Closed before 2016")
  ) +
  tm_facets(by=c("System","Year"))+
  tm_credits(c("52 schools\n(18,771 OTGC)","49 schools\n(18,039 OTGC)",
                "95 schools\n(41,796 OTGC)","88 schools\n(40,249 OTGC)"),
             position = c("LEFT", "BOTTOM")) +
  tmap_defaults +
  tm_layout(bg.color = "grey85",
            panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),
                                c('2011','2016')),
            panel.label.bg.color = "white")

catchment_schools_plot
```

```{r descriptive-catchment-schools-plot2, eval=FALSE}
tmap_save(catchment_schools_plot, paste0(here::here(),"/Figures/Fig1.tiff"), height=7, width=12, dpi=600)
```

```{r,include=FALSE}
load(file = paste0(here::here(),"/data-prep/data-products/avail_2011.RData"))
load(file = paste0(here::here(),"/data-prep/data-products/avail_2016.RData"))
```

@@

```{r, eval=FALSE}
#check to make sure that V_i is equal to the total capacity in Hamilton. It is!
avail_2011$V_ij %>% sum()
ALL_SCHOOLS_Elem$OTGC2011 %>% sum(na.rm = TRUE)

avail_2016$V_ij %>% sum()
ALL_SCHOOLS_Elem$OTGC2016 %>% sum(na.rm = TRUE)

@@

#this should equal 58,865, still does -- great!
avail_2016 %>%
  group_by(ID) %>%
  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()
```

@@

#add the geometries -- census DAs
SAvail_im_DA2011 <- SAvail_im_DA2011 |>
  left_join(cen_data_2011_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

SAvail_im_DA2016 <- SAvail_im_DA2016 |>
  left_join(cen_data_2016_hamiltononly |> filter(Region.Name == "Hamilton") |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

#Create 1 sf with data joined (no changes, just 2011 and 2016)
SAvail_im_DA <-
  rbind(SAvail_im_DA2011 |> mutate(year ="2011"),
        SAvail_im_DA2016 |> mutate(year ="2016")) |>
  mutate(mode = case_when(mode == "c" ~ "motorized",
                          mode == "w" ~ "non-motorized"))
```

```{r}
#calc changes between the years - but at the 2016 DA level 
SAvail_im_DA2011_forGeoUID2016 <- avail_2011 %>%
  group_by(GeoUID_2016, 
           mode) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj),
            V_im = sum(V_ij),
            kid_popadj_m = sum(kid_popadj),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
            .groups = "drop") |>
  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
         v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT))

>>>>

---
title: "10 years of school closures and consolidations: the impact on school accessibility"
bibliography: 
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
format:
  nature-pdf: 
    journal: "default"
    keep-tex: true
editor: source
execute: 
  freeze: auto

author:
  - name: "Anastasia Soukhov"
    email: "soukhoa@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
    attributes:
      corresponding: true
  - name: "Christopher D. Higgins"
    email: "cd.higgins@utoronto.ca"
    affiliations:
      - id: 2
        address: "Department of Human Geography, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON, M1C 1A4, Canada"
  - name: "Antonio Páez"
    email: "paezha@mcmaster.ca"
    affiliations:
      - id: 1
        address: "School of Earth, Environment and Society, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
  - name: "Moataz Mohamed"
    email: "mmohame@mcmaster.ca"
    affiliations:
      - id: 3
        address: "Department of Civil Engineering, McMaster University, 1241 Main St. West, Hamilton, ON, L8S 4K1, Canada"
abstract: | 
  | Reducing motorized travel and increasing active travel are top priorities for many communities. However, these priorities are confronted with pressures to reduce the fiscal burden of public services, including schools. Public school boards have been strained to do more with less, and opting to close under-capacity schools to deliver educational services more cost-effectively. However, once a school is closed, students' proximity to schools decreases, and those who previously lived close enough to walk or cycle can no longer do so. In this paper, the City of Hamilton, a mid-sized city in Ontario, Canada, which experienced a 10% decrease in elementary school locations between 2011 and 2021, serves as a case study. We quanitfiy how these closures and consolidations have reshaped accessibility for the elementary school-aged population and discuss their implications for active and motorized travel.  To measure accessibility, spatial availability, a singly-constrained multimodal measure of spatial accessibility is used. This method's proportional allocation mechanism enables interpretable comparison between accessibility 'before' and 'after' the implementation of school closure/consolidation policy considering changes in potential travel time and population. The analysis is conducted at the parcel level, considering all school seats as eligible and incorporating observed home-to-school travel patterns for motorized and non-motorized modes. We present system-wide impact estimates resulting from these closures. The findings reveal that overall spatial availability in the city declined. While the school consolidation policy may have achieved cost savings in facility operation and maintenance, the burden falls on students, families, and communities through _potentially_ reduced opportunities for physical activity through an active journey to school and increased motorized travel externalities.

keywords: [accessibility; spatial availability; carbon emissions; journey-to-school; walkability; active transportation; school consolidation policy; service provision thresholds]
---
<!--NOTE WHEN PREPARING THE .TEX for journal submission I ran into errors as their system uses PDFlatex+Bibtex to compile. The following packages only work in lualatex, so do the following:
- remove "  \usepackage[T1]{fontenc}" 
- remove "\usepackage{fontspec}" 
- add "\DeclareUnicodeCharacter{394}{$\Delta$}" to the tex preamble (this is because PHDtex does not recognize unicode? So need to explicitly define) (add to line 5) -->

<!--Chat with Antonio about the assumption regarding the 'student population' per parcel. As it stands, when the student pop for the OD matrix is summed, it equals the total student population in the city. Each parcel is assigned a DA-pop/(the number of potential parcel trips in a DA * the potential parcels trips from that parcel). This is highly correlated with the value just weighted by DA-pop/DA-dwellings (but need to double check).  --->

@@

```{r, include=FALSE}
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggspatial) # Spatial Data Framework for ggplot2
library(here) # A Simpler Way to Find Your Files
library(patchwork) # The Composer of Plots
library(scales) # Scale Functions for Visualization
library(sf) # Simple Features for R
library(stplanr) # Sustainable Transport Planning
library(tidyverse) # Easily Install and Load the 'Tidyverse'
library(tmap) # Thematic Maps
library(units) # Measurement Units for R Vectors
library(miscTools) # Miscellaneous Tools and Utilities
# library(huxtable)
library(tiff)
library(grid)
library(cowplot)
library(flextable)
library(gt)
```

@@

ham_city_bound <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/City_Boundary.shp"),
                          quiet = TRUE) |> st_transform(st_crs(cen_data_2016_hamiltononly))

#Add % urbanization to the label for ham_community_bounds
# SAvail_im_DA <- left_join(SAvail_im_DA, DA_2021_URBvsRUR, by="GeoUID")
# 
# SAvail_im_DA |> filter(mode == "Motorized") |> st_drop_geometry() |> group_by(COMMUNITY) |> 
#   summarise(perc_urb = mean(perc_urb, na.rm=TRUE))

ham_community_bounds <- st_read(paste0(here::here(),"/data-prep/data-inputs/Boundaries/Community_Boundaries.shp"),quiet = TRUE) |> 
  st_transform(st_crs(cen_data_2016_hamiltononly)) |> 
  mutate(COMMUNITY = case_when(COMMUNITY_ == "Ancaster" ~ "AN",
                               COMMUNITY_ == "Dundas" ~ "DU",
                               COMMUNITY_ == "Flamborough" ~ "FL",
                               COMMUNITY_ == "Glanbrook" ~ "GL",
                               COMMUNITY_ == "Hamilton" ~ "HA",
                               COMMUNITY_ == "Stoney Creek" ~ "SC")) |> st_transform(st_crs(cen_data_2016_hamiltononly)) |> 
  mutate(COMMUNITY_URB = case_when(COMMUNITY == "AN" ~ "Ancaster (87%)",
                               COMMUNITY == "DU" ~ "Dundas (96%)",
                               COMMUNITY == "FL" ~ "Flamborough (37%)",
                               COMMUNITY == "GL" ~ "Glanbrook (47%)",
                               COMMUNITY == "HA" ~ "Hamilton Central (100%)" ,
                               COMMUNITY == "SC" ~"Stoney Creek (94%)"),
         COMMUNITY_ = ifelse(COMMUNITY=="HA", "Hamilton Central", COMMUNITY_))

@@

cen_data_2021_hamiltononly <- 
  cen_data_2021_hamiltononly %>% dplyr::select( #removing the variables we don't need for this analysis
  -c("Shape Area",  "Type", "CD_UID", "CSD_UID", "CT_UID", "CT_UID", "v_CA21_566: Median after-tax income in 2020 among recipients ($)" ,"v_CA21_71: 15 to 19 years")) %>%
  rename( #renaming the variables
    "DA_households" = "Households",
    "DA_dwellings" = "Dwellings",
    "DA_tot_pop" = "Population",
    "DA_med_AT_family_income" = "v_CA21_966: Median after-tax income of economic family in 2020 ($)",
    "DA_med_AT_lone-family_income" = "v_CA21_978: Median after-tax income of one-parent economic families in 2020 ($)",
    "DA_prev_LIMAT_under18" =  "v_CA21_1043: 0 to 17 years",
    "DA_pop_5to9" = "v_CA21_32: 5 to 9 years",
    "DA_pop_10to14" = "v_CA21_50: 10 to 14 years") %>%
    mutate(DA_pop_5to14 = DA_pop_5to9+(DA_pop_10to14)) |> #*0.9)) %>%
         dplyr::select(-c("DA_pop_5to9","DA_pop_10to14"))

#adding $ URBAN to the 2021 census DA file... retrieved from parcel points. No other year of parcel points provide this information 
load(paste0(here::here(),file="/data-prep/data-products/DA_2021_URBvsRUR.Rdata"))
cen_data_2021_hamiltononly <- cen_data_2021_hamiltononly |> left_join(DA_2021_URBvsRUR, by="GeoUID")
```

@@

```{r formatting-schools-for-mapping, include=FALSE}
#reformatting schools for mapping
ALL_SCHOOLS_Elem_formapping_11 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
  mutate(combo_status = "1) No change",
         combo_year = "2011",
         OTGC = OTGC2011) 
ALL_SCHOOLS_Elem_formapping_12 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
  mutate(combo_status = "1) No change",
         combo_year = "2016",
         OTGC = OTGC2016) 
ALL_SCHOOLS_Elem_formapping_13 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "NoChange" & Status_2 == "NoChange") |>
  mutate(combo_status = "1) No change",
         combo_year = "2021",
         OTGC = OTGC2021)

ALL_SCHOOLS_Elem_formapping_21 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
  mutate(combo_status = "2) Expanded between 2011-16",
         combo_year = "2011",
         OTGC = OTGC2011)
ALL_SCHOOLS_Elem_formapping_22 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
  mutate(combo_status = "2) Expanded between 2011-16",
         combo_year = "2016",
         OTGC = OTGC2016)
ALL_SCHOOLS_Elem_formapping_23 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016 and 2021" & Status == "Expanded" & Status_2 == "NoChange") |>
  mutate(combo_status = "2) Expanded between 2011-16",
         combo_year = "2021",
         OTGC = OTGC2021)

ALL_SCHOOLS_Elem_formapping_31 <- SCHOOLS |>
  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
  mutate(combo_status = "3) Relocated + expanded between 2011-16",
         combo_year = "2016",
         OTGC = OTGC2016)
ALL_SCHOOLS_Elem_formapping_32 <- SCHOOLS |>
  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "Moved" & Status_2 == "NoChange") |>
  mutate(combo_status = "3) Relocated + expanded between 2011-16",
         combo_year = "2021",
         OTGC = OTGC2021)

ALL_SCHOOLS_Elem_formapping_41 <- SCHOOLS |>
  filter(Year == "2011" & is.na(Year_2) & Status == "Moved" & is.na(Status_2)) |>
  mutate(combo_status = "3) Relocated + expanded between 2011-16",
         combo_year = "2011",
         OTGC = OTGC2011)

ALL_SCHOOLS_Elem_formapping_51 <- SCHOOLS |>
  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
  mutate(combo_status = "6) Opened between 2011-16",
         combo_year = "2016",
         OTGC = OTGC2016) 
ALL_SCHOOLS_Elem_formapping_52 <- SCHOOLS |>
  filter(Year == "2016" & Year_2 == "2016 and 2021" & Status == "New" & Status_2 == "NoChange") |>
  mutate(combo_status = "6) Opened between 2011-16",
         combo_year = "2021",
         OTGC = OTGC2021)

ALL_SCHOOLS_Elem_formapping_81 <- SCHOOLS |>
  filter(is.na(Year) & Year_2 == "2021" & is.na(Status) & Status_2 == "New") |>
  mutate(combo_status = "7) Opened between 2016-21",
         combo_year = "2021",
         OTGC = OTGC2021)

ALL_SCHOOLS_Elem_formapping_71 <- SCHOOLS |>
  filter(Year == "2011" & is.na(Year_2) & Status == "Removed" & is.na(Status_2)) |>
  mutate(combo_status = "4) Closed between 2011-16",
         combo_year = "2011",
         OTGC = OTGC2011)

ALL_SCHOOLS_Elem_formapping_61 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
  mutate(combo_status = "5) Closed between 2016-21",
         combo_year = "2011",
         OTGC = OTGC2011)

ALL_SCHOOLS_Elem_formapping_62 <- SCHOOLS |>
  filter(Year == "2011 and 2016" & Year_2 == "2016" & Status == "NoChange" & Status_2 == "Removed") |>
  mutate(combo_status = "5) Closed between 2016-21",
         combo_year = "2016",
         OTGC = OTGC2016)

ALL_SCHOOLS_Elem_formapping <- rbind(
  ALL_SCHOOLS_Elem_formapping_11,ALL_SCHOOLS_Elem_formapping_12, ALL_SCHOOLS_Elem_formapping_13,
  ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,ALL_SCHOOLS_Elem_formapping_23, 
  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32, 
  ALL_SCHOOLS_Elem_formapping_41,
  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
  ALL_SCHOOLS_Elem_formapping_71,
  ALL_SCHOOLS_Elem_formapping_81
  )

rm(ALL_SCHOOLS_Elem_formapping_11,ALL_SCHOOLS_Elem_formapping_12, ALL_SCHOOLS_Elem_formapping_13,
  ALL_SCHOOLS_Elem_formapping_21,ALL_SCHOOLS_Elem_formapping_22,ALL_SCHOOLS_Elem_formapping_23, 
  ALL_SCHOOLS_Elem_formapping_31,ALL_SCHOOLS_Elem_formapping_32, 
  ALL_SCHOOLS_Elem_formapping_41,
  ALL_SCHOOLS_Elem_formapping_51,ALL_SCHOOLS_Elem_formapping_52,
  ALL_SCHOOLS_Elem_formapping_61,ALL_SCHOOLS_Elem_formapping_62,
  ALL_SCHOOLS_Elem_formapping_71,
  ALL_SCHOOLS_Elem_formapping_81)
```

```{r descriptive-schools-plot, eval=FALSE}
tmap_defaults <-
  tm_scale_bar(position = c("LEFT", "BOTTOM"), breaks = c(0, 5, 10, 15))+
  tm_compass(size = 1.5, position = c("right", 'top')) +
  tm_layout(legend.position = c("left", "top")) #legend.hist.size = 10,

schools_plot <-
  tm_shape(ham_bay, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=ham_community_bounds) + tm_polygons(col="lightblue", border.alpha = 0)+

  tm_shape(cen_data_2021_hamiltononly |> filter(!is.na(perc_urb))) +
  tm_polygons(col = "perc_urb", border.alpha=0, 
              alph=0.8,
              palette = c("yellow3","yellow4"),
              #labels = c("[0-50%)", "[25-50%)", "[50-75%)","[75-100%]"),
              breaks = c(0,  0.25, .5, .75,  1),
              legend.is.portrait = FALSE,
              style="cont",
              title = "Proportion urban land-use") +

  tm_shape(ALL_SCHOOLS_Elem_formapping) +
  tm_symbols(shape = "combo_status",
             col="grey85",
             size= "OTGC",
             alpha=0.9,
             scale = 0.4,
             shapes = c(1,
                        17,24,
                        15,22,
                        18, 23),
          title.shape = "School status",
          title.size = "OTGC",
          shapes.labels =  c(
          "No change (btwn 2011-21)", 
          "Expanded (btwn 2011-16)", 
          "Relocated + expanded (btwn 2011-16)",
          "Closed (btwn 2011-16)", 
          "Closed (btwn 2016-21)",
          "Opened (btwn 2011-16)", 
          "Opened (btwn 2016-21)")) +
  tm_facets(by=c("System","combo_year"), nrow = 3)+

  tm_credits(c("52 schools open\n(18,771 OTGC)","49 schools open\n(18,039 OTGC)", "49 schools open\n(18,218 OTGC)",
                "95 schools open\n(41,796 OTGC)","88 schools open\n(40,249 OTGC)","83 schools open\n(39,478 OTGC)"),
             position = c("LEFT", "BOTTOM")) +
  tm_shape(ham_city_bound)+  tm_borders(col = "grey80", lwd=0.5)+
  tmap_defaults +
  tm_layout(bg.color = "grey99",
            panel.labels = list(c('HWDCSB (Catholic Public)','HWDSB (Public)'),
                                c('2011','2016','2021')),
            panel.label.bg.color = "white")

tmap_save(schools_plot, paste0(here::here(),"/Figures/Fig2.tiff"), height=4.5, width=10, dpi=600)
```

```{r,include=FALSE}
load(file = paste0(here::here(),"/data-prep/data-products/avail_2011.RData"))
load(file = paste0(here::here(),"/data-prep/data-products/avail_2016.RData"))
load(file = paste0(here::here(),"/data-prep/data-products/avail_2021.RData"))
```

@@

```{r, eval=FALSE}
#check to make sure that V_i is equal to the total capacity in Hamilton. It is!
avail_2011$V_ij %>% sum()
SCHOOLS$OTGC2011 %>% sum(na.rm = TRUE)

avail_2016$V_ij %>% sum()
SCHOOLS$OTGC2016 %>% sum(na.rm = TRUE)

avail_2021$V_ij %>% sum()
SCHOOLS$OTGC2021 %>% sum(na.rm = TRUE)

@@

#this should equal 58,865, still does -- great!
avail_2016 %>%
  group_by(ID) %>%
  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()

#this should equal 62195, still does -- great!
avail_2021 %>%
  group_by(ID) %>%
  summarize(kid_popadj = sum(kid_popadj)) |> dplyr::select(kid_popadj) %>% sum()
```

@@

#calc spatial availability and other associated variables at the DA 2021 level
SAvail_im_DA2021 <- avail_2021 %>%
  group_by(GeoUID, 
           mode) %>%
  summarize(COMMUNITY = first(COMMUNITY),
            total_tt = sum(travel_time_p50*kid_popadj),
            V_im = sum(V_ij),
            kid_popadj_m = sum(kid_popadj),
            DA_pop_5to14 = mean(DA_pop_5to14, na.rm=T),
            DA_prev_LIMAT_under18 = mean(DA_prev_LIMAT_under18),
            .groups = "drop") |>
  mutate(emissions_total_car_tt = ifelse(mode == "c", 0.1*total_tt, 0),
         v_im_kid_popadj = V_im/kid_popadj_m,
         v_im_kid_popadj = ifelse(is.na(v_im_kid_popadj) | is.nan(v_im_kid_popadj), 0, v_im_kid_popadj),
         v_im_kid_popadj_LIMAT = v_im_kid_popadj*(DA_prev_LIMAT_under18/100),
         v_im_kid_popadj_LIMAT = ifelse(is.na(v_im_kid_popadj_LIMAT) | is.nan(v_im_kid_popadj_LIMAT), 0, v_im_kid_popadj_LIMAT),
         V_i_kid_popadj_LIMAT = V_im*(DA_prev_LIMAT_under18/100),
         V_i_kid_popadj_LIMAT = ifelse(is.na(V_i_kid_popadj_LIMAT) | is.nan(V_i_kid_popadj_LIMAT), 0, V_i_kid_popadj_LIMAT)) |>
  filter(GeoUID != 35250211 & GeoUID != 35250242 & GeoUID != 35250491 & GeoUID != 35250982 ) #these GeoUID's have 0 kid population, so let's drop so we don't have zeros
```

```{r}
#add the geometries -- census DAs
SAvail_im_DA2011 <- SAvail_im_DA2011 |>
  left_join(cen_data_2011_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

SAvail_im_DA2016 <- SAvail_im_DA2016 |>
  left_join(cen_data_2016_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

SAvail_im_DA2021 <- SAvail_im_DA2021 |>
  left_join(cen_data_2021_hamiltononly |> dplyr::select(GeoUID), by="GeoUID") |> st_set_geometry("geometry")

#Create 1 sf with data joined (no changes, just 2011 and 2016)
SAvail_im_DA <-
  rbind(SAvail_im_DA2011 |> mutate(year ="2011"),
        SAvail_im_DA2016 |> mutate(year ="2016"),
        SAvail_im_DA2021 |> mutate(year ="2021")) |>
  mutate(mode = case_when(mode == "c" ~ "Motorized",
                          mode == "w" ~ "Non-motorized"))
```

```{r}
#calc changes between the years - but do it all at the 2021 DA grid. So let's calculate all SAvails at the level of the 2021 grid.
